---
title: Calculate sample size or power for published microbiome studies with IMPACTT
  guidelines
author: "David Xin Zhao"
date: "Last edited `r format(Sys.time(), '%d %B %Y')`"
knit: (function(inputFile, encoding) {
      out_dir <- 'docs';
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_file=file.path(dirname(inputFile), out_dir, 'index.html'))})
output:
  html_document:
    # theme: cosmo
    highlight: pygments
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    collapsed: FALSE
    number_sections: TRUE
    fig_width: 7
    fig_height: 6
    fig_caption: TRUE
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

<html>

<head>

```{=html}
<style>

h1{
 color: #055C9D;
 font-family: Georgia;
 font-size: 200%
}


h2{
 color: #055C9D;
 font-family: helvetica;
 font-size: 150%
}

h3{
 color: #055C9D;  
 font-family: helvetica;
 font-size: 120%; 
}

p {
 color: #333333;
 font-family: helvetica;
 font-size: 100%;
}

.blackbox {
  padding: 1em;
  background: green;
  color: black;
  border: 2px solid orange;
  border-radius: 10px;
}

.center {
  text-align: center;
}

</style>
```
</head>

</html>

```{r setup, include = FALSE}
# set options for the entire document 
knitr::opts_chunk$set(fig.align = 'center', 
                      fig.height=6, fig.width=8,
                      dev="png",
                      echo=TRUE, #display code in output document 
                      error=FALSE,
                      collapse = FALSE, 
                      message=FALSE) #stop render when error occurs   
```

## Problem

Calculating the sample size appropriately is critical in planning a
valid microbiome study. Insufficient sample size will cause poor power
in a hypothesis testing while more-than-enough sample size will cause
resource wasting or even animal/ human ethics problem.

Power is the probability to correctly reject a null hypothesis when it
is false, which indicates how credible a microbiome study is.

Microbiome data has special characteristics different from common data
set, which thus require special methods to calculate sample size or
power.

## Solution

IMPACTT experts provided detailed guidelines in 2021-2022
[@ferdous2022][@beta-div2022], helping microbiome researchers calculate
sample size and power appropriately.

## Project outline

Sample size required is specific to types of hypothesis to test. Four
types of hypothesis testing are covered in the project as follows,

1.  Worked example: beta-diversity: IPA study [@chen2021]
2.  Worked example: alpha-diversity: IPA study [@chen2021]
3.  Worked example: differential abundance analysis of a taxon: vitamin
    D study [@zhao2023]
4.  Worked example: microbial cluster membership: food sensitization
    [@tun2021]
5.  Worked example: taxon membership (colonization with a taxon): C.
    difficile colonization [@drall2020]; B. longum subsp. infantis
    colonization [@chen2021]

## R codes

In this project, I applied the effective framework to real dataset,
while practicing R coding skills and familiarizing statistical concepts.

### Case study: sample size or power for comparing within-group and between-group distance matrix (beta-diversity)

```{r beta-distance}

# select appropriate beta-diversity measurement 



## distribution assumption: normal distribution 

# method 1: Equation A in Table 2 

# method 2: ANOVA F-test R-square (Equation G in Table 2) 




## distribution assumption: nonparametric distribution 

# scenario 1: distance matrix is known 

# scenario 2: distance matrix is unknown - simulate distance matrix  




## distribution assumption: Dirichlet-Multinomial distribution (La Rosa et al.) 

# method: HMP R package  


```

### Case study: Alpha-diversity

Alpha diversity refers to the number of species present in an ecosystem
(richness) as well as the frequency of occurrence of each type of
organisms (evenness).

Commonly used metrics include:

-   Shannon

-   Inverse Simpson

-   Simpson

-   Chao indices

-   Faith's phylogenetic diversity (Faith's PD)

```{r alpha-diversity}


## distribution assumption: normal distribution 

# method 1: Equation A  (explanatory variable is binary) 

# method 2: Equation G for correlation (explanatory variable is continuous)


# convert medians and interquartile ranges to means and standard deviations  

```

### Case study: Differential analysis of taxon abundance

Two types of hypothesis tests:

-   Compare mean abundance of a taxon

-   Compare subject proportion with abundance over a chosen threshold

Type I error threshold, alpha should not need adjustment for multiple
testing, if testing on only single taxon.

Type I error threshold, alpha should be adjusted for multiple testing,
if a study plans to test association at all taxa.

alpha\^star = alpha / M (where M is the planned number of tests)

```{r taxon abundance}

## distribution assumption: ?

# Equation D and E in Table 2 



## distribution assumption: non-normal distribution 

# convert median and IQR to mean and standard deviation (suitable for phylum or higher classification levels) 


```

### Case study: Testing higher or lower rates of cluster membership between groups

```{r cluster membership}

# Equation C an E in table 2 


```

### Case study: Testing higher or lower rates of taxon membership between groups (i.e. colonization with a microbe)

To determine sample sizes for colonization with specific microbiota
(presence/ absence or yes/ no), one can use the equations for
proportions or odds ratios.

```{r taxa colonization}

# Equation C in table 2 



```

### Case study: Longitudinal study (SIMR: An R package for power analysis of generalized linear mixed models by simulation)
